---
title: "What Affects People's Intention to Have Children in Canada"
author: "Sheng Kai, Shiqi Liu, Wuxin Dong"
date: '2020-10-19'
bibliography: "references.bib"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Authors: Sheng Kai, Shiqi Liu, Wuxin Dong 
# 2020-10-19
# Code and data supporting this analysis is available at: https://github.com/EricD0406/304PS-2


## Abstract

As the trend of population aging increases, the fertility intentions of families have become a more crucial topic for the society as a whole. In this paper, we explore the survey data collected by the 2017 General Social Survey (GSS) on the Family to better understand what factors will affect people's intention for having children and how these factors will affect their intentions in Canada. Our findings have shown that the gender, age, health condition and education level will all affect people's decisions about whether or not to have children. Our study and analysis could have implications for policy makers and family researchers who are interested in people's fertility intention, since our study results could help them to better revise policy to encourage fertility and conduct future researches respectively.

## Introduction

In today's society, the structures of family have changed a lot, but the role family plays in people's lives is irreplaceable. Children have always been an important component of the family, because people's intention to have children could imply the characteristics and socio-economic conditions of families. Understanding what affects people's intention to have children is an interesting topic for family researchers and policy makers. Many researches about whether or not people want to have children have been conducted, whose results can help them measure the changes in living conditions and well-being of families, and to assess the effects of family-related policy. Therefore, to determine what factors will affect people's intention to have children, and the number of total children they are intending to have is of significance.

Individual attitudes and behaviors regarding the intentions to have children tend to be influenced by their family system, communication and the attitudes of other family members (Marisa & Anne, 2017). Besides, personal attitudes towards life will also affect people's decision-making process on whether or not to have children, which means people who have positive attitude through the life are more likely to intend to have children, or to have more children than those who are relatively pessimistic. Furthermore, fertility intentions seem to differ between different gender groups. More specifically, men are more likely to have intentions for children than women. Moreover, educational level also contributes to people's decision-making about fertility. Some research has shown that people with higher education levels are less likely to intend to have children (Maria, 2014).

In this paper, through the study of the 2017 General Social Survey (GSS) on the Family, we find that people's feelings about life, their satisfaction towards life will affect people's intention to have children. Both physical and mental self-health conditions matter when deciding to have children. Beyond that, there are splits between different age and gender groups in terms of their fertility intentions.

For analysis, we will firstly do data-cleaning to choose and preserve variables of our interest, which include number_total_children_intention, age, sex, education, self_rated_health, and self_rated_mental_health. To further investigate the demographic distribution of respondents, we also include income_respondent and region. With regard to the model, we would like to use Logistic Regression to analyze the effect of above factors on people's intention for whether or not having children.

```{r, echo = FALSE, message=FALSE}
#install.packages("tidyverse")
#install.packages("dplyr")
#install.packages("kableExtra")
#install.packages("survey")
library(tidyverse)
library(dplyr)
library(kableExtra)
library(survey)
fig_cap <- captioner::captioner()
tab_cap <- captioner::captioner("Table")

```


## Data

The data used in the following analysis was obtained from the 2017 General Social Survey (GSS) on the Family and cleaned using an R Package created by Alexander and Caetano (2020). Two primary objectives of the GSS are to gather data on social trends to monitor changes in the living conditions and well-being of Canadians over time, and to provide information on specific social policy. 

### Key Features

The 2017 GSS focused on families' characteristics, diversity and future. Two comprised components of the data are core content and classification variables, of which the core content is to measure changes in well-being, and effects of specific policy issues, and classification variables (such as, age, sex, education and income) are to collect demographic information for use in the analysis of core data.

### Methodology

The data set consists of both numerical variables and categorical variables which exhibits the demographics and characteristics of families in Canada. The target population for the 2017 GSS included all persons 15 years of age older in Canada, excluding residents of the Yukon, Northwest Territories, and Nunavut, and full-time residents of institutions. Regarding the probability sampling method, they use stratified random sampling by dividing 10 provinces into strata, which are geographic areas. With many of the Census Metropolitan Areas (CMAs) being considered as separate strata, there were 27 strata in total. Using stratified random sampling can reduce the bias and variability of results, because within one strata, additional samples are homogeneous, thus statistics for sampled values will vary less across samples.

The sampling frame of the data was a combination of the list of telephone number (both landline and cellular) in use available to Statistics Canada and the Address Register (AR), which is the list of all dwellings within the 10 provinces. They use all telephone numbers (linked and not linked to the AR), which could make sure a coverage of all households with telephones. They also included cellular phone number this year, which enlarged the frame by adding households without landlines. Because many young families might not have landlines nowadays.

Within each geographic stratum, they carry out a simple random sampling without replacement to obtain the sample. The target sample size was 20,000, while the actual number of respondents was 20,602. The sample size was large, and was proportionally chosen from 10 provinces, therefore, the sample was representative of the target population. However, households without telephones or not registered on AR were just excluded from the sampled population. This would omit new families who had not had the chance to register and the families with poor economic conditions who could not afford the telephone bills.

Data collection for the 2017 GSS was conducted via computer assisted telephone interviews (CATI) in the official language of respondents' choice. The interview calls were made from approximately 9 AM to 9:30 PM on workdays, with schedule extension from 10 AM to 5 PM on Saturdays and 1 PM to 9 PM on Sundays. The extension of interview schedule made it more flexible for respondents, which could help increase the response rate. However, calls were only made during the daytime on weekdays, and it was very likely that there was no one at home for working families during those times. This could cause an obvious decline of response rate.

For non-response, they re-contacted up to two more times to explain the importance of the survey to those who refused to participate at first. They also arranged appointment to call back at a more convenient time for those who were inconvenient at the first call. For cases in which there was no one at home, numerous call backs were made. The combination of these strategies dealing with 3 different situations would increase the response rate. However, the associated costs were high due to large time and human resources consumption. Besides, the re-arrangement of the interview could not be very efficient, because respondents might be unavailable again during the newly-scheduled time, or they might forget it.

### Data Type and Selection

In our analysis, we will focus on the variable number_total_children_intention, age, sex, income_respondent, education, self_rated_health, and self_rated_mental_health. Since we would like to investigate the factors influencing people's intention for having children, we will use number_total_children_intention as our response variable in the model. Another similar variable in the dataset is future_children_intention, which is a categorical variable. We decided to choose number_total_children_intention as our response variable, because the answers for the question collecting data for future_children_intention are different categories, "Probably yes" and "Probably no" are too ambiguous for analysis. We are not sure respondents' true intention for children by referring these category anwers. While number_total_children_intention is a numerical variable with certain numerical answers, which is easier for quantity analysis, and to be transformed to a binary variable.


### Questionnaire for Data Collection

For variable age, it is confirmed by asking the respondents to provide their date of birth. With this variable provided, some questions later in the survey are then determined to be asked or not. It enhances the accuracy of this variable and responses related to this variable. Nevertheless, since the survey was conducted through a phone call, it becomes more difficult to verify the information.

For variable income, no questions would be asked, and the information would be obtained through a linkage to tax data. This will increase the quality of this variable as it eliminate the burden of respondents. One thing to notice is that there were people who rejected the linkage and no linkage to their tax data took place.

For variable self_rated_health and self_rated_mental_health, the question was "In general, would you say your (mental) health is...?". They provided a range for respondents to make it more convenient for them to answer. However, the questions might be too general, because they did not clearly explain what each category means. For example, 'Poor" in terms of mental health could mean that you always feel unhappy and disordered and you feel hopeless about life.

For variable number_total_children_intention, it collected people's intentions about giving birthing or having more children. Moreover, questions about what contraceptive methods they use are also asked. Therefore, the result should be of high quality and give a clearer picture of fertility. However, there were more than half of repsondents just skipped this question.

### Raw Data Overview


```{r, echo = FALSE}
gss <- read.csv("gss.csv")
#Raw Data Table
raw_data <- gss%>%select(number_total_children_intention, age, sex,
                      education, self_rated_health,
                      self_rated_mental_health, income_respondent)


sex <- raw_data%>%group_by(sex)%>%count()%>%
  mutate(prop=paste(round(n/nrow(raw_data),2)*100,"%"))
education <- raw_data%>%group_by(education)%>%count()%>%
  mutate(prop=paste(round(n/nrow(raw_data),2)*100,"%"))
self_rated_mental_health <- raw_data%>%group_by(self_rated_mental_health)%>%count()%>%
  mutate(prop=paste(round(n/nrow(raw_data),2)*100,"%"))
self_rated_health <- raw_data%>%group_by(self_rated_health)%>%count()%>%
  mutate(prop=paste(round(n/nrow(raw_data),2)*100,"%"))
income_respondent <- raw_data%>%group_by(income_respondent)%>%count()%>%
  mutate(prop=paste(round(n/nrow(raw_data),2)*100,"%"))


demogr1 <- rbind(sex,
                education%>%rename(sex=education),
                self_rated_mental_health%>%rename(sex=self_rated_mental_health),
                self_rated_health%>%rename(sex=self_rated_health),
                income_respondent%>%rename(sex=income_respondent))
colnames(demogr1) <- c("Variable", "Count", "Proportion")

demogr1 <- demogr1%>%kbl(caption="Respondent Demographics",
                       align=c("lrr"))%>%
pack_rows("Sex",1,2)%>%pack_rows("Education",3,10)%>%pack_rows("Self Rated Mental Health",11,17)%>%pack_rows("Self Rated Health",18,24)%>%pack_rows("Income Respondent",25,30)
demogr1
```
`r tab_cap(1,"Raw Data Overview")`

As shown in table 1, it gives people a brief overview of our selected raw data. Since the report will review the results of the clean version of this table which will be table 3 in our result part, we will focus on what changes we need to make to our raw data. We can see that there are NAs in Education and Self rated health. In addition, there are "Don't know" in the self rated mental health and self rated health. Therefore, we would consider to remove these unknowns as to make our analysis more useful. Furthermore, we can see from education that there are so many categories under it and we may need to group some of them together to make our report cleaner and more representative. Last but not the least, number_total_children_intention and age are not included in this table because they are numeric variables. As a result, we decided to clean our data by putting age into different 5 age groups and number_total_children_intention into 2 groups with yes and no. 


## Model

```{r, echo = FALSE}
#Variables of interest
raw_data <- gss%>%select(number_total_children_intention, age, sex,
                      education, self_rated_health,
                      self_rated_mental_health, income_respondent)
#Rename
mdata <- raw_data%>%rename(health = self_rated_health,
                        mental_health=self_rated_mental_health,
                        income=income_respondent)
#Filter NA
mdata <- mdata%>%filter(!is.na(number_total_children_intention))
mdata <- mdata%>%filter(mental_health != "Don't know")
mdata <- mdata%>%filter(health != "Don't know")
mdata <- mdata%>%filter(!is.na(education))
#Group age
mdata$age_group = mdata$age
mdata$age_group[mdata$age_group<= 25] <- "15-25"
mdata$age_group[mdata$age_group<= 35 & mdata$age_group > 25] <- "25-35"
mdata$age_group[mdata$age_group<= 45 & mdata$age_group > 35] <- "35-45"
mdata$age_group[mdata$age_group<= 55 & mdata$age_group > 45] <- "45-55"
mdata$age_group[mdata$age_group > 55] <- "55-80"
#Group Education
mdata$education[mdata$education == 'High school diploma or a high school equivalency certificate'] <- 'High School'
mdata$education[mdata$education == "Bachelor's degree (e.g. B.A., B.Sc., LL.B.)"] <- 'Bachelor'
mdata$education[mdata$education == "University certificate, diploma or degree above the bach..."] <- 'Bachelor'
mdata$education[mdata$education == "College, CEGEP or other non-university certificate or di..."] <- 'College'
mdata$education[mdata$education == "University certificate or diploma below the bachelor's level"] <- 'College'
mdata$education[mdata$education == "Less than high school diploma or its equivalent"] <- 'Below High School'
mdata$education[mdata$education == "Trade certificate or diploma"] <- 'Trade Cert'
#Children intention
mdata<-mdata%>%rename(children_intention = number_total_children_intention)
mdata$intentionYN = mdata$children_intention
mdata$intentionYN[mdata$intentionYN != 0] <- "Yes"
mdata$intentionYN[mdata$intentionYN == 0] <- "No"
mdata$children_intention[mdata$children_intention != 0] <- 1
mdata$children_intention[mdata$children_intention == 0] <- 0
```
### Data Cleaning

As mentioned before, we will focus on the variable number_total_children_intention, age, sex, income_respondent, education, self_rated_health, and self_rated_mental_health. We cleaned our data by renaming self_rated_health as health, self_rated_mental_health as mental_health, income_respondent as income and number_total_children_intention as children_intention. Then, we have removed all NAs in the children_intention and education as they were meaningless for our later analysis. For the same reason, we have also decided to remove all “Don’t know” values in the variables health and mental_health. 
  
In addition, we divided age into 5 groups, which were 15 - 25, 25 -35, 35-45, 45-55 and above 55. The reason we did that is that we would like to look at the percentage in each age group later in our table 1: Respondent Demographic. After that, we have also renamed and reduced the categories of the degrees as to make our later analysis clearer. The details about how we renamed and reduced the categories of the degrees under education will be in our appendix^1. In the end, since the majority of the responses for the question "Total number of children intending to have" are 0, we decide to divide the responses into two categories "1" and "0". Responses greater than 0 will be redefined as "1", while answers with 0 correspond to "0". Therefore, by running the logistic regression, we would like to find what affects people's decision on whether or not to have children. This binary variable with "1" representing people having fertility intention, while 0 representing people with no intention for having children.

We divided respondents to different age groups when exhibiting the demographic table because this will makes it more convenient to identify the distribution of respondents' ages at the aggregate level. However, in our logistic regression model, we still use age rather than age groups, since we would like to see if there is a trend that people's intentions for children will decrease as age increases. Therefore, using age would be more helpful for our analysis. Besides, gender and education level, health conditions are all categorical variables, because these variables' values represent different categories of respondents. Among these categorical variables, except gender, all other 3 are ordinal categorical variables, because they represent different levels in which respondents are.

### Model building and selection

We selected children_intention as the response variable, which represents whether people want to have children or not, while the candidate predictor variables include age, sex, self rated health, self rated mental health and education. Since children intention is a binary response variable that has two possible outcomes, a logistic regression model was fitted to estimate the probability that someone intend to have children using all predictor variables. The reason why linear regression model is not appropriate here is that it will cause the response variable to exceed 1 or drop below 0, while probability must be between 0 and 1. Logistic regression model works better in predicting binary response variable as it take the form of log($\frac{p}{1-p}$) = $\beta_0$ + $\beta_1X_1$ + $\beta_2X_2$ + $\cdots$ + $\beta_kX_k$ In current study, p represents the probability of intention to have children, Xs represent the value of predictor variables, and  $\beta$s are the coefficients that we want to derive from the fitted model.
Also, as we are using a survey data, we applied finite population correction to adjust the variance estimate for our estimated mean probability of intention to have children. The target population size is  30302287 by estimation, and the sample size after data cleaning is 8339. 
In terms of software, we have used Rstudio to run out model.
According to the first model (table 2) fitted using all candidate predictors, self rated health was not statistically significant, suggesting that people rated their health condition in different levels won't affect their intention to have children. Therefore, we removed self rated health from the predictors, and the final model only include age, sex, self rated mental health and education as predictor variables.


```{r, echo = FALSE}

N <- 30302287
n <- nrow(mdata)
fpc.srs = rep(N, n)

intention.design <- svydesign(id=~1, data=mdata, fpc=fpc.srs)

mysvylm <- svyglm(children_intention ~ age+sex+
             health+mental_health + education, intention.design,family="binomial")
#summary(mysvylm)

```

Coefficients |      Estimate     | Std. Error | t value | Pr(>|t|) 
-----------|:-----------: |:-----------: |:-----------: |:-----------: |
(Intercept)                |6.735434    |0.181971  |37.014  |< 2e-16 ***
age                        |-0.197509   |0.004705  |-41.976 |< 2e-16 ***
sexMale                    |0.678505    |0.059234  |11.455  |< 2e-16 ***
healthFair                 |0.135754    |0.163250  |0.832   |0.405675    
healthGood                 |0.059717    |0.089997  |0.664   |0.507006    
healthPoor                 |-0.179565   |0.273441  |-0.657  |0.511400    
healthVery good            |-0.005551   |0.078291  |-0.071  |0.943480    
mental_healthFair          |-0.983360   |0.145937  |-6.738  |1.71e-11 ***
mental_healthGood          |-0.437811   |0.088014  |-4.974  |6.68e-07 ***
mental_healthPoor          |-1.116372   |0.266125  |-4.195  |2.76e-05 ***
mental_healthVery good     |-0.280481   |0.077857  |-3.603  |0.000317 ***
educationBelow High School |-1.295755   |0.148857  |-8.705  |< 2e-16 ***
educationCollege           |-0.258151   |0.069528  |-3.713  |0.000206 ***
educationHigh School       |-0.704451   |0.086347  |-8.158  |3.90e-16 ***
educationTrade Cert        |-0.448523   |0.116699  |-3.843  |0.000122 ***

`r tab_cap(2,"First Model with age, sex, health, mental_health, education as predictors")`

## Result

The results will be presented in 3 sections according to the following characteristics: numeric results from logistic model, demographic profile and children intention under each variable.

### Numeric Results from Logistic Model

```{r, echo = FALSE}
mysvylm2 <- svyglm(children_intention ~ age+sex+
          mental_health + education, intention.design,family="binomial")
#summary(mysvylm2)

```
Coefficients |      Estimate     | Std. Error | t value | Pr(>|t|) 
-----------|:-----------: |:-----------: |:-----------: |:-----------: |
(Intercept)                |6.729232    |0.180621  |37.256  |< 2e-16 ***
age                        |-0.197160   |0.004684  |-42.088 |< 2e-16 ***
sexMale                    |0.679359    |0.059191  |11.477  |< 2e-16 ***
mental_healthFair          |-0.946465   |0.135755  |-6.972  |3.37e-12 ***
mental_healthGood          |-0.410005   |0.076675  |-5.347  |9.17e-08 ***
mental_healthPoor          |-1.107626   |0.247633  |-4.473  |7.82e-06 ***
mental_healthVery good     |-0.279409   |0.070868  |-3.943  |8.13e-05 ***
educationBelow High School |-1.284758   |0.148362  |-8.660  |< 2e-16 ***
educationCollege           |-0.256493   |0.069484  |-3.691  |0.000224 ***
educationHigh School       |-0.698500   |0.085857  |-8.136  |4.70e-16 ***
educationTrade Cert        |-0.445445   |0.116397  |-3.827  |0.000131 ***

`r tab_cap(3,"Final Model with age, sex, mental_health, education as predictors")`

As shown in table 3 above, for every unit increase in age, we expect the log odds of intention of having children to decrease by 0.197. When the respondent is a male, then we expect the log odds of intention of having children to increase by 0.679. When the mental health situation are poor, fair, good, very good or excellent, we expect the log odds of intention of having children to decrease by 1.107, 0.946, 0.41, 0.279 or 0 respectively. When the education level is below high school, high school, trade cert, college or bachelor, we expect the log odds of intention of having children to decrease by 1.284,  0.699, 0.445, 0.257.

### Demographic Profile

```{r, echo = FALSE}
#Demographic table
sex <- mdata%>%group_by(sex)%>%count()%>%
  mutate(prop=paste(round(n/nrow(mdata),2)*100,"%"))
age_group <- mdata%>%group_by(age_group)%>%count()%>%
  mutate(prop=paste(round(n/nrow(mdata),2)*100,"%"))
education <- mdata%>%group_by(education)%>%count()%>%
  mutate(prop=paste(round(n/nrow(mdata),2)*100,"%"))
mental_health <- mdata%>%group_by(mental_health)%>%count()%>%
  mutate(prop=paste(round(n/nrow(mdata),2)*100,"%"))


demogr <- rbind(sex,age_group%>%rename(sex=age_group),
                education%>%rename(sex=education),
                mental_health%>%rename(sex=mental_health))
colnames(demogr) <- c("Variable", "Count", "Proportion")
demogr <- demogr%>%kbl(caption="Respondent Demographics",
                       align=c("lrr"))%>%
  pack_rows("Sex",1,2)%>%pack_rows("Age",3,7)%>%pack_rows("Education",8,12)%>%pack_rows("Mental_Health",13,17)
demogr
```
`r tab_cap(4,"Respondent Demographic")`

As shown in table 4, 52% female and 48% male constituted our selected respondents. For Age, the breakdown consisted of 17% between 15 and 25, 30%  between 25 - 35, 35% between 35 - 45, 18% of 45 - 55 and 1 % above 55. For education, the breakdowns included 34% bachelor's degree and above, 9% below high school degree, 28% college degree, 22% high school degree and 7% trade certificates. For mental health, the responses consisted of 29% excellent, 7% fair, 27% good, 2% poor and 35% very good.


### Children Intention under Each Variable

```{r, echo = FALSE}
##Children intention
intention <- mdata%>%group_by(children_intention)%>%count()
x <-  intention$n
labels <-  intention$children_intention
piepercent<- round(100*x/sum(x), 1)

pie(x, labels = piepercent, main = " Percentage of Intention to Have Children in Canada",col = rainbow(length(x)))
legend("topright", c("No","Yes"), cex = 0.8,
   fill = rainbow(length(x)))
```

`r fig_cap(1,"Chilren Intention")`

As shown in figure 1, about 59% of the respondents have no intention to have a kid and 41 % of the respondents have intention to have one or more kids.


```{r, echo = FALSE}
#Children intention (group by sex)
mdata %>% 
    group_by(sex) %>% 
    count(intentionYN) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = sex, y = prop)) +
    geom_col(aes(fill = intentionYN), position = "dodge") +
    geom_text(aes(label = scales::percent(prop), 
                  y = prop, 
                  group = intentionYN),
              position = position_dodge(width = 0.9),
              vjust = 1.5)
```

`r fig_cap(2,"Children Intention by sex")`

As shown in figure 2, female respondents have a 63.3% intention of not having a child. Male respondents have a 54.6% intention of not having a child. 

```{r, echo = FALSE}
#Children intention (group by education)
mdata %>% 
    group_by(education) %>% 
    count(intentionYN) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = education, y = prop)) +
    geom_col(aes(fill = intentionYN), position = "dodge") +
    geom_text(aes(label = scales::percent(prop), 
                  y = prop, 
                  group = intentionYN),
              position = position_dodge(width = 0.9),
              vjust = 1.5)
```

`r fig_cap(3,"Children Intention by education")`

As shown in figure 3, 60.8% of respondents with bachelor degree have no fertility intentions, 42% of respondents with below high school degree have no fertility intentions, 64.2% of respondents with college degree have no fertility intentions, 56.1 % of respondents with high school degree have no fertility intentions and 62.2% of respondents with trade certificates no fertility intentions. 



```{r, echo = FALSE}
#Children intention (group by mental health)
mdata %>% 
    group_by(mental_health) %>% 
    count(intentionYN) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = mental_health, y = prop)) +
    geom_col(aes(fill = intentionYN), position = "dodge") +
    geom_text(aes(label = scales::percent(prop), 
                  y = prop, 
                  group = intentionYN),
              position = position_dodge(width = 0.9),
              vjust = 1.5)
```

`r fig_cap(4,"Children Intention by mental health")`

As shown in figure 4, 53.4% of respondents with excellent mental health have no fertility intentions, 65.2% of respondents with fair mental health have no fertility intentions, 62.8 % of respondents with good mental health have no fertility intentions, 66.4% of respondents with poor mental health have no fertility intentions and 59.5% of respondents with very good mental health have no fertility intentions. 

```{r, echo = FALSE}
#Children intention (group by age group)
mdata %>% 
    group_by(age_group) %>% 
    count(intentionYN) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = age_group, y = prop)) +
    geom_col(aes(fill = intentionYN), position = "dodge") +
    geom_text(aes(label = scales::percent(prop), 
                  y = prop, 
                  group = intentionYN),
              position = position_dodge(width = 0.9),
              vjust = 1.5)
```

`r fig_cap(5,"Children Intention by age_group")`

As shown in figure 5, 16.4% of respondents of age between 15 and 25 have no fertility intentions, 36.1% of respondents of age between 25 and 35 have no fertility intentions, 79.4% of respondents of age between 35 and 45 have no fertility intentions, 95.2 % of respondents of age between 45 and 55 have no fertility intentions and 94.1% of respondents of age between 55 and 88 have no fertility intentions. 



